\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-11-16 submission/manuscript.tex
Coauthors review done.

\usepackage{lineno}
\renewcommand{\linenumberfont}{\normalfont\tiny}

@@

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,071
different OTUs with 382 to 2,031 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,686.1 ±
142.8 OTUs{]} and monospecific {[}Number of OTUs, 1,729.5 ± 159.6
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,064.1 ± 210.0
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 525.4 ± 145.1 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (41.4 -- 52.4 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.30, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 52.4
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

Sampling was performed in the Bay of Funtana, northern Adriatic Sea
(\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in
a meadow of \emph{Cymodocea nodosa} invaded by the invasive
\emph{Caulerpa cylindracea} (mixed settlement) and in a monospecific
settlement of \emph{Caulerpa cylindracea} located in the proximity of
the meadow at approximately monthly intervals from December 2017 to
October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were
collected by diving and transported to the laboratory in containers
placed on ice and filled with seawater collected at the sampling site.
Upon arrival to the laboratory, \emph{C. nodosa} leaves were cut into
sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea} thalli were cut
into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three
times with sterile artificial seawater (ASW) to remove loosely attached
microbial cells. Ambient seawater was collected in 10 \si{\l} containers
by diving and transported to the laboratory where the whole container
volume was filtered through a 20 \si{\um} net. The filtrate was further
sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate
membrane filters (Whatman, United Kingdom) using a peristaltic pump.
Filters were briefly dried at room temperature and stored at \num{-80}
\si{\degreeCelsius}. Seawater samples were also collected approximately
monthly from July 2017 to October 2018.

@@

This work was founded by the Croatian Science Foundation through the
MICRO-SEAGRASS project (IP-2016-06-7118). We would like to thank
Margareta Buterer for technical support, Paolo Paliaga for help during
sampling and the University Computing Center of the University of Zagreb
for access to the computer cluster Isabella.

@@

\leavevmode\hypertarget{ref-Korlevica}{}%
Korlević, M., Markovski, M., Zhao, Z., Herndl, G.J., and Najdek, M.
Selective DNA and protein isolation from marine macrophyte surfaces.

@@

\caption{Taxonomic classification and relative contribution of the most abundant (> 1 \si{\percent}) bacterial and archaeal sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. NR -- No Relative (sequences without known relatives within the corresponding group)\label{community}}\label{fig:unnamed-chunk-4}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant (> 1 \si{\percent}) cyanobacterial sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of cyanobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{cyano}}\label{fig:unnamed-chunk-5}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant (> 2 \si{\percent}) sequences within the \textit{Bacteroidota} on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \textit{Bacteroidota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{bactero}}\label{fig:unnamed-chunk-6}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant (> 2 \si{\percent}) alphaproteobacterial sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of alphaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{alpha}}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant (> 2 \si{\percent}) gammaproteobacterial sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of gammaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{gamma}}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant (> 1 \si{\percent}) sequences within the \textit{Desulfobacterota} on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \textit{Desulfobacterota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{desulfo}}\label{fig:unnamed-chunk-9}
\end{figure}

>>>>

%\usepackage{lineno}
%\renewcommand{\linenumberfont}{\normalfont\tiny}

@@

\sisetup{mode=text}
\setlength\parindent{24pt}

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,082
different OTUs with 379 to 2,017 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,698.7 ±
139.6 OTUs{]} and monospecific {[}Number of OTUs, 1,735.2 ± 160.1
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,052.8 ± 216.1
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 528.8 ± 145.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (37.8 -- 53.6 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 53.6
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

@@

Sampling was performed in the Bay of Funtana, northern Adriatic Sea
(\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in
a meadow of \emph{Cymodocea nodosa} invaded by the invasive
\emph{Caulerpa cylindracea} (mixed settlement) and in a monospecific
settlement of \emph{Caulerpa cylindracea} located in the proximity of
the meadow at approximately monthly intervals from December 2017 to
October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were
collected by diving and transported to the laboratory in containers
placed on ice and filled with seawater collected at the sampling site.
Upon arrival to the laboratory, \emph{C. nodosa} leaves were cut into
sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea} thalli were cut
into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three
times with sterile artificial seawater (ASW) to remove loosely attached
microbial cells. Ambient seawater was collected in 10 \si{\l} containers
by diving and transported to the laboratory where 10 -- 20 \si{\l} were
filtered through a 20 \si{\um} net. The filtrate was further
sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate
membrane filters (Whatman, United Kingdom) using a peristaltic pump.
Filters were briefly dried at room temperature and stored at \num{-80}
\si{\degreeCelsius}. Seawater samples were also collected approximately
monthly from July 2017 to October 2018.

@@

This work was founded by the Croatian Science Foundation through the
MICRO-SEAGRASS project (project number IP-2016-06-7118). ZZ and GJH were
supported by the Austrian Science Fund (FWF) project ARTEMIS (project
number P28781-B21). We would like to thank Margareta Buterer for
technical support, Paolo Paliaga for help during sampling and the
University Computing Center of the University of Zagreb for access to
the computer cluster Isabella.

@@

\leavevmode\hypertarget{ref-Korlevic}{}%
Korlević, M., Markovski, M., Zhao, Z., Herndl, G.J., and Najdek, M.
Selective DNA and protein isolation from marine macrophyte surfaces.

@@

\caption{Taxonomic classification and relative contribution of the most abundant ($\geq$ 1 \si{\percent}) bacterial and archaeal sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. NR -- No Relative (sequences without known relatives within the corresponding group)\label{community}}\label{fig:unnamed-chunk-4}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant ($\geq$ 1 \si{\percent}) cyanobacterial sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of cyanobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{cyano}}\label{fig:unnamed-chunk-5}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant ($\geq$ 2 \si{\percent}) sequences within the \textit{Bacteroidota} on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \textit{Bacteroidota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{bactero}}\label{fig:unnamed-chunk-6}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant ($\geq$ 2 \si{\percent}) alphaproteobacterial sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of alphaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{alpha}}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant ($\geq$ 1 \si{\percent}) gammaproteobacterial sequences on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of gammaproteobacterial sequences in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{gamma}}\label{fig:unnamed-chunk-8}
\end{figure}

@@

\caption{Taxonomic classification and relative contribution of the most abundant ($\geq$ 1 \si{\percent}) sequences within the \textit{Desulfobacterota} on the surfaces of the macrophytes \textit{C. nodosa} (mixed settlement) and \textit{C. cylindracea} (mixed and monospecific settlement) and in the ambient seawater. The proportion of sequences classified as \textit{Desulfobacterota} in the total bacterial and archaeal community is given above the corresponding bar. NR -- No Relative (sequences without known relatives within the corresponding group)\label{desulfo}}\label{fig:unnamed-chunk-9}
\end{figure}
